Diffantom: a software layer to build digital phantoms of whole-brain diffusion MRI

### 2015-08-12 paper/main.tex
I hope this fixes #30

\section*{Introduction}
Fiber tracking on \gls*{dmri} data has become an important application for the in-vivo investigation
  of the structural configuration of fiber bundles at the macro-scale.
The final outcome of such applications, called tractography, is fundamental to gain information about
  \gls*{wm} morphology in many clinical routines like neurosurgical planning \citep{golby_interactive_2011},
  post-surgery evaluations \citep{toda_utility_2014},
  and the study of neurological diseases as in \citep{chua_diffusion_2008} addressing multiple sclerosis and
  Alzheimer's disease.
Tractography is also crucial in the analysis of structural brain networks using graph theory.
These analyses have been successfully applied in the definition of the unique subject-wise patterns of connectivity
  \citep{sporns_human_2005}, the assessment of neurological diseases \citep{griffa_structural_2013}, and in the
  study of the link between structural and functional connectivity \citep{messe_predicting_2015}.
However, the development of the field is limited by the lack of a gold standard to test and compare the
  wide range of methodologies available for processing and analyzing \gls*{dmri}.
In this report we present \emph{diffantom}, an \emph{in-silico} dataset to assess tractography pipelines using
  real data as source of whole-brain microstructural information.
We also provide a full software layer to produce \emph{diffantoms} from real \gls*{dmri} datasets
  drawn from the \gls*{hcp}.
The \emph{diffantom} enables the following evaluations of connectivity pipelines:
\begin{itemize}
  \item Studies on the impact of different diffusion sampling schemes on the local micro-structure
  model of choice and on the subsequent global tractography outcome.
  \item Assessment of sensitivity and robustness to imaging artifacts (noise, \acrlong{pve} and \glslines{csf} contamination,
    susceptibility-derived warping, Eddy-currents-derived distortions, etc.) at both local and global levels.
  \item Characterizing the \acrfull*{roc} of connectivity pipelines.
  \item Generation of synthetic connectivity networks to test graph theory analyses.
  \item Simulation of pathological brains (e.g. tumors as in \cite{kaus_simulation_2000}).
\end{itemize}

\section*{Background}
The assessment of tractography from \gls*{dmri} data is challenging due to
  the lack of a gold standard.
Large efforts have been devoted to the development of physical phantoms
  \citep{lin_validation_2001,campbell_flowbased_2005,perrin_validation_2005,fieremans_simulation_2008,tournier_resolving_2008}.
\cite{cote_tractometer_2013} conducted a thorough review of tractography methodologies using the
  so-called \emph{FiberCup} phantom \citep{poupon_new_2008,fillard_quantitative_2011}.
These phantoms are appropriate to evaluate the angular resolution in fiber crossings and accuracy of
  direction-independent scalar parameters in very simplistic geometries.
Digital simulations are increasingly popular because the complexity of whole-brain tractography
  can not be accounted for with current materials and proposed methodologies to build physical phantoms.
Early digital phantoms also started with simulation of simple geometries
  \citep{basser_in_2000,goessl_fiber_2002,tournier_limitations_2002,leemans_mathematical_2005}
  to evaluate resolution of fiber crossings.
These tools generally implemented the multi-tensor model \citep{alexander_analysis_2001,tuch_high_2002}
  to simulate fiber crossing, fanning, kissing, etc.
\cite{close_software_2009} presented the \emph{Numerical Fibre Generator}, a software to simulate
  spherical shapes filled with digital fiber-tracts.
\cite{caruyer_phantomas_2014} proposed \emph{Phantomas} to simulate any kind of analytic geometry
  inside a sphere.
\emph{Phantomas} models diffusion by a restricted and a hindered compartment, similar to
  \citep{assaf_composite_2005}.
\cite{wilkins_fiber_2014} proposed a whole-brain simulated phantom derived from voxel-wise orientation
  of fibers averaged from real \gls*{dmri} scans and the multi-tensor model with a compartment of
  isotropic diffusion.
Recently, \cite{neher_fiberfox_2014} proposed \emph{FiberFox}, a visualization software to develop
  complex geometries and their analytical description.
Once the geometries are obtained, the software generates the corresponding \gls*{dmri} signal with a
  methodology very close to that implemented in \emph{Phantomas}.

@@

\noindent\textbf{\textit{Data Generation\textcolon}\label{sec:data_generation}}
The methods to generate \emph{diffantoms} are rooted in the work of \cite{wilkins_fiber_2014}, and
  described in \autoref{fig:figure1}.
\emph{Diffantom} integrates two major improvements with respect to \citep{wilkins_fiber_2014}.
First, since we use a dataset from the \gls*{hcp} as input, data are corrected for the most relevant distortions.
\cite{wilkins_fiber_2014} explicitly state that their original data were not subjected to certain distortion
  corrections, and thus, generated data are affected correspondingly.
Second improvement is a more advanced signal model to generate the phantom using
  \emph{Phantomas} \citep{caruyer_phantomas_2014}.

The simulation process is started by generating a microstructural model from real data (\autoref{fig:figure1}).
On one hand, the software requires up to five volume fractions $\{T_{1\,\cdots\,5}\}$ corresponding to the free-diffusion
  compartments that will be simulated.
These isotropic fraction maps use the \emph{5TT format} established with the latest version 3.0 of
  \emph{MRTrix} \citep{tournier_mrtrix_2012}, that describes the structure of the brain in the following
  order: \gls*{cgm}, \gls*{dgm}, \gls*{wm}, \gls*{csf}, and abnormal tissue.
Since we simulate a healthy subject, the last fraction map is empty and can be omitted.
On the other hand, the hindered-diffusion compartments are specified by up to three volume fractions $\{F_{1\,\cdots\,3}\}$
  of three single fiber populations per voxel along with their corresponding direction maps $\{V_{1\,\cdots\,3}\}$.

@@

Once a microstructural model of the volume has been synthesized, a map of \glspl{fod} is computed using spherical
  convolution of the single fiber response and the fiber orientation maps $\{V_{1\,\cdots\,3}\}$, weighted by the
  corresponding fiber-fractions $\{F_{1\,\cdots\,3}\}$.
A close-up showing how the \glspl{fod} map looks is presented in \autoref{fig:figure1}.
The single fiber response is a Gaussian diffusion tensor with axial symmetry and eigenvalues $\lambda_1=$ 2.2\e{-3}
  and $\lambda_{2,3}=$ 0.2\e{-3} [mm$^2$s$^{-1}$].
The resulting \glspl{fod} map is then combined with the isotropic compartments corresponding to $\{T_{1\,\cdots\,5}\}$
  before generating the signal using \emph{Phantomas} \citep{caruyer_phantomas_2014}.
By default, diffusion data are generated using a scheme of 100 directions distributed in one shell with uniform
  coverage \citep{caruyer_design_2013}.
Custom one- or multi-shell schemes can be generated supplying the tables of corresponding $b$-vectors and $b$-values.
Rician noise is also included in \emph{Phantomas}, and can be set by the user.
The default value for \gls*{snr} is preset to 90.0.
\Gls*{csf} was simulated with a unique compartment with isotropic diffusivity $D_{CSF}$ of 3.0\e{-3} [mm$^2$s$^{-1}$].
Conversely, \Gls*{wm} and \gls*{gm} included one isotropic compartment with $D_{WM} =$ 2.0\e{-4}, $D_{cGM} =$ 7.0\e{-4}
  and $D_{dGM} =$ 9.0\e{-4}, respectively [mm$^2$s$^{-1}$].
All these values for diffusivity (and the corresponding to the single-fiber response) can be modified by the user with
  custom settings.

\noindent\textbf{\textit{Interpretation and use\textcolon}}
The full simulation workflow, presented in \autoref{fig:figure1}, is implemented using
  \emph{nipype} \citep{gorgolewski_nipype_2011} to ensure reproducibility and usability.
The recommended use case of the simulation workflow is its integration in assessment frameworks
  (\autoref{fig:figure2}, box A) for the evaluation of algorithms and/or pipelines.

To illustrate the features of the \emph{diffantom}, the example dataset underwent a simplified
  connectivity pipeline including \gls*{csd} and probabilistic tractography from
  \emph{MRTrix} \citep{tournier_mrtrix_2012}.
\Gls*{csd} was configured with up to 8 spherical harmonics, and tractography with 1.6$\times$10$^\text{6}$
  seed points evenly distributed across a dilated mask of the \gls*{wm} tissue.
\autoref{fig:figure2}, box B1 shows the result of the tractography obtained with such pipeline.
Finally, we applied \emph{tract querier} \citep{wassermann_on_2013} on segmenting important fiber bundles such
  as the \gls*{cst} and the forceps minor (see \autoref{fig:figure2}, box B2).
Particularly, due to its location nearby the orbitofrontal lobe, the forceps minor is generally affected by
  susceptibility distortions.
The study of this phenomena \citep{esteban_simulationbased_2014} led us to design \emph{diffantom}.

\section*{Discussion and conclusion}

\noindent\textbf{\textit{Discussion\textcolon}}\label{sec:discussion} %
Whole-brain, realistic \gls*{dmri} phantoms are necessary in the developing field of structural
  connectomics.
The \emph{diffantom} is a derivative of \citep{wilkins_fiber_2014} in terms of methodology for
  simulation with two major advances: the correctness of the \emph{minimally preprocessed} data
  \citep{glasser_minimal_2013} released within the \gls*{hcp}, and a more realistic diffusion
  model.
A possible competitor to \emph{diffantom} is the ground-truth data generated by the organizers of the
  Tractography Challenge held in the ISMRM 2015\footnote{\url{http://www.tractometer.org/ismrm_2015_challenge/}}.
Similarly to \emph{diffantom}, one \gls*{hcp} dataset is used as source of structural information.
However, in this case 25 fiber bundles were manually segmented to obtain a summary tractogram.
Then, making use of \emph{FiberFox}, the segmentation of each bundle is mapped to an analytical description
  which is fed into \emph{Phantomas}.
While this phantom is designed for the bundle-wise evaluation of tractography (with scores such as geometrical coverage,
  valid connections, invalid connections, missed connections, etc. \citep{cote_tractometer_2013}),
  \emph{diffantom} is intended for the connectome-wise evaluation of results, yielding tractography with
  a large number of bundles.
Therefore, \emph{diffantom} and \emph{FiberFox} are complementary as the hypotheses that can be investigated are different.
Moreover, \emph{diffantom} does not require costly manual segmentation of bundles, highly demanding in terms of physiology
  expertise and operation time.
This facilitates the use of \emph{diffantom} as a factory of whole-brain diffusion phantoms.
Finally, since the gradient scheme can be set by the user, \emph{diffantom} can be seen as a mean to translate the so-called
  \emph{b-matrix} of the \gls*{hcp} to any target scheme.

>>>>

\section*{Introduction}
Fiber tracking on \gls*{dmri} data has become an important application for the in-vivo investigation
  of the structural configuration of fiber bundles at the macro-scale.
The final outcome of such applications, called tractography, is fundamental to gain information about
  \gls*{wm} morphology in many clinical routines like neurosurgical planning \citep{golby_interactive_2011},
  post-surgery evaluations \citep{toda_utility_2014},
  and the study of neurological diseases as in \citep{chua_diffusion_2008} addressing multiple sclerosis and
  Alzheimer's disease.
Tractography is also crucial in the analysis of structural brain networks using graph theory.
These analyses have been successfully applied in the definition of the unique subject-wise patterns of connectivity
  \citep{sporns_human_2005}, the assessment of neurological diseases \citep{griffa_structural_2013}, and in the
  study of the link between structural and functional connectivity \citep{messe_predicting_2015}.
However, the development of the field is limited by the lack of a gold standard to test and compare the
  wide range of methodologies available for processing and analyzing \gls*{dmri}.

Large efforts have been devoted to the development of physical phantoms
  \citep{lin_validation_2001,campbell_flowbased_2005,perrin_validation_2005,fieremans_simulation_2008,tournier_resolving_2008}.
\cite{cote_tractometer_2013} conducted a thorough review of tractography methodologies using the
  so-called \emph{FiberCup} phantom \citep{poupon_new_2008,fillard_quantitative_2011}.
These phantoms are appropriate to evaluate the angular resolution in fiber crossings and accuracy of
  direction-independent scalar parameters in very simplistic geometries.
Digital simulations are increasingly popular because the complexity of whole-brain tractography
  can not be accounted for with current materials and proposed methodologies to build physical phantoms.
Early digital phantoms also started with simulation of simple geometries
  \citep{basser_in_2000,goessl_fiber_2002,tournier_limitations_2002,leemans_mathematical_2005}
  to evaluate resolution of fiber crossings.
These tools generally implemented the multi-tensor model \citep{alexander_analysis_2001,tuch_high_2002}
  to simulate fiber crossing, fanning, kissing, etc.
\cite{close_software_2009} presented the \emph{Numerical Fibre Generator}, a software to simulate
  spherical shapes filled with digital fiber-tracts.
\cite{caruyer_phantomas_2014} proposed \emph{Phantomas} to simulate any kind of analytic geometry
  inside a sphere.
\emph{Phantomas} models diffusion by a restricted and a hindered compartment, similar to
  \citep{assaf_composite_2005}.
\cite{wilkins_fiber_2014} proposed a whole-brain simulated phantom derived from voxel-wise orientation
  of fibers averaged from real \gls*{dmri} scans and the multi-tensor model with a compartment of
  isotropic diffusion.
Recently, \cite{neher_fiberfox_2014} proposed \emph{FiberFox}, a visualization software to develop
  complex geometries and their analytical description.
Once the geometries are obtained, the software generates the corresponding \gls*{dmri} signal with a
  methodology very close to that implemented in \emph{Phantomas}.

\emph{Diffantom} is originally designed for the investigation of susceptibility-derived distortions, a
  typical artifact that produces geometrical warping in certain regions of \gls*{dmri} datasets.
In \citep{esteban_simulationbased_2014} we addressed this phenomenon and concluded that the connectivity
  matrix of \emph{Phantomas} was not dense enough to evaluate the integration of correction methods
  in pipelines for the connectome extraction.
In this report we present \emph{diffantom}, an \emph{in-silico} dataset to assess pipelines for tractography
  and connectome extraction using real data as source of whole-brain microstructural information.
\emph{Diffantom} is inspired by the work of \cite{wilkins_fiber_2014}, with two principal novelties.
First, since we use a dataset from the \gls*{hcp} as input, data are corrected for the most relevant distortions.
\cite{wilkins_fiber_2014} explicitly state that their original data were not subjected to certain distortion
  corrections, and thus, generated data are affected correspondingly.
The second improvement is a more advanced signal model to generate the phantom using the hindered and restricted
  diffusion model of \emph{Phantomas} \citep{caruyer_phantomas_2014}.

@@

\noindent\textbf{\textit{Microstructural model\textcolon}\label{sec:data_model}} %
The simulation process relies on a microstructural model derived from real data (\autoref{fig:figure1}).
On one hand, the software requires up to five volume fractions $\{T_{1\,\cdots\,5}\}$ corresponding to the
  free-diffusion compartments.
These fraction maps are obtained from the macroscopic structure of tissues in the brain,
  in the following order~%
\footnote{Corresponding to the \emph{5TT format} established with the latest version 3.0
  of \emph{MRTrix} \citep{tournier_mrtrix_2012}}:
\gls*{cgm}, \gls*{dgm}, \gls*{wm}, \gls*{csf}, and abnormal tissue~%
\footnote{Since here we simulate healthy subjects, the last fraction map is empty and can be omitted}.
On the other hand, the hindered-diffusion compartments are specified by up to three volume fractions $\{F_{1\,\cdots\,3}\}$
  of three single fiber populations per voxel along with their corresponding direction maps $\{V_{1\,\cdots\,3}\}$.

@@

\noindent\textbf{\textit{Diffusion signal generation\textcolon}\label{sec:data_dwi}} %
Once a microstructural model of the volume has been synthesized, a map of \glspl{fod} is computed using spherical
  convolution of the single fiber response and the fiber orientation maps $\{V_{1\,\cdots\,3}\}$, weighted by the
  corresponding fiber-fractions $\{F_{1\,\cdots\,3}\}$.
A close-up showing how the \glspl{fod} map looks is presented in \autoref{fig:figure1}.
The single fiber response is a Gaussian diffusion tensor with axial symmetry and eigenvalues $\lambda_1=$ 2.2\e{-3}
  and $\lambda_{2,3}=$ 0.2\e{-3} [mm$^2$s$^{-1}$].
The resulting \glspl{fod} map is then combined with the isotropic compartments corresponding to $\{T_{1\,\cdots\,5}\}$
  before generating the signal using \emph{Phantomas} \citep{caruyer_phantomas_2014}.
By default, diffusion data are generated using a scheme of 100 directions distributed in one shell with uniform
  coverage \citep{caruyer_design_2013}.
Custom one- or multi-shell schemes can be generated supplying the tables of corresponding $b$-vectors and $b$-values.
Rician noise is also included in \emph{Phantomas}, and can be set by the user.
The default value for \gls*{snr} is preset to 90.0.
\Gls*{csf} was simulated with a unique compartment with isotropic diffusivity $D_{CSF}$ of 3.0\e{-3} [mm$^2$s$^{-1}$].
Conversely, \Gls*{wm} and \gls*{gm} included one isotropic compartment with $D_{WM} =$ 2.0\e{-4}, $D_{cGM} =$ 7.0\e{-4}
  and $D_{dGM} =$ 9.0\e{-4}, respectively [mm$^2$s$^{-1}$].
All these values for diffusivity (and the corresponding to the single-fiber response) can be modified by the user with
  custom settings.

\noindent\textbf{\textit{Implementation and reproducibility\textcolon}\label{sec:data_workflow}} %
We also provide a full software layer to produce \emph{diffantoms} from real \gls*{dmri} datasets
  drawn from the \gls*{hcp}.
This workflow, presented in \autoref{fig:figure1}, is implemented using
  \emph{nipype} \citep{gorgolewski_nipype_2011} to ensure reproducibility and usability.

\noindent\textbf{\textit{Interpretation and recommended uses\textcolon}\label{sec:data_use}} %
To illustrate the features of the \emph{diffantom}, the example dataset underwent a simplified
  connectivity pipeline including \gls*{csd} and probabilistic tractography from
  \emph{MRTrix} \citep{tournier_mrtrix_2012}.
\Gls*{csd} was configured with up to 8 spherical harmonics, and tractography with 1.6$\times$10$^\text{6}$
  seed points evenly distributed across a dilated mask of the \gls*{wm} tissue.
\autoref{fig:figure2}, box B1 shows the result of the tractography obtained with such pipeline.
Finally, we applied \emph{tract querier} \citep{wassermann_on_2013} on segmenting important fiber bundles such
  as the \gls*{cst} and the forceps minor (see \autoref{fig:figure2}, box B2).
Particularly, due to its location nearby the orbitofrontal lobe, the forceps minor is generally affected by
  susceptibility distortions.

Since the gradient scheme can be set by the user, \emph{diffantom} can be seen as a mean to translate the so-called
  \emph{b-matrix} of the \gls*{hcp} to any target scheme.
A second recommended use case of the simulation workflow is its integration in assessment frameworks
  (\autoref{fig:figure2}, box A) for the unit testing of algorithms, the integration tests of
  modules in workflows and the full validation of pipelines.
The \emph{diffantom} enables the following evaluations of connectivity pipelines:
\begin{itemize}
  \item Studies on the impact of different diffusion sampling schemes on the local micro-structure
  model of choice and on the subsequent global tractography outcome.
  \item Assessment of sensitivity and robustness to imaging artifacts (noise, \acrlong{pve} and \glslines{csf} contamination,
    susceptibility-derived warping, Eddy-currents-derived distortions, etc.) at both local and global levels.
  \item Characterizing the \acrfull*{roc} of connectivity pipelines.
  \item Generation of synthetic connectivity networks to test graph theory analyses.
  \item Simulation of pathological brains (e.g. tumors as in \cite{kaus_simulation_2000}).
\end{itemize}

\section*{Discussion and conclusion}
\noindent\textbf{\textit{Discussion\textcolon}}\label{sec:discussion} %
Whole-brain, realistic \gls*{dmri} phantoms are necessary in the developing field of structural
  connectomics.
The \emph{diffantom} is a derivative of \citep{wilkins_fiber_2014} in terms of methodology for
  simulation with two major advances: the correctness of the \emph{minimally preprocessed} data
  \citep{glasser_minimal_2013} released within the \gls*{hcp}, and a more realistic diffusion
  model.
A possible competitor to \emph{diffantom} is the ground-truth data generated by the organizers of the
  Tractography Challenge held in the ISMRM 2015\footnote{\url{http://www.tractometer.org/ismrm_2015_challenge/}}.
Similarly to \emph{diffantom}, one \gls*{hcp} dataset is used as source of structural information.
However, in this case 25 fiber bundles were manually segmented to obtain a summary tractogram.
Then, making use of \emph{FiberFox}, the segmentation of each bundle is mapped to an analytical description
  which is fed into \emph{Phantomas}.
While this phantom is designed for the bundle-wise evaluation of tractography (with scores such as geometrical coverage,
  valid connections, invalid connections, missed connections, etc. \citep{cote_tractometer_2013}),
  \emph{diffantom} is intended for the connectome-wise evaluation of results, yielding tractography with
  a large number of bundles.
Therefore, \emph{diffantom} and \emph{FiberFox} are complementary as the hypotheses that can be investigated are different.
Moreover, \emph{diffantom} does not require costly manual segmentation of bundles, highly demanding in terms of physiology
  expertise and operation time.
This facilitates the use of \emph{diffantom} as a factory of whole-brain diffusion phantoms.
